---
title: "Time series *plastics* with Holt-Winters algorithms"
author: "Matthew O'Callaghan"
date: " 2019 "
output: 
   html_document: default
   
---
<!-- ----------------------------------------------- --> 

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!-- ----------------------------------------------- --> 
<!-- ----------------DATA IMPORT-------------------- --> 
<!-- ----------------------------------------------- --> 
## Plastics Data:
According to page 39 of the fma package [documentation ](www.https://cloud.r-project.org/web/packages/fma/fma.pdf) we have the following information about the date we are using throughout this assignment: 

* **Description**
  + Monthly sales of product A for a plastics  manufacturer.
* **Usage**
  + plastics
* **Format**
  + Time series data
  
  
```{r message=FALSE}
# Importing forecasting package where plastics is contained:
require("fma")
require("ggplot2")
plastics
```
The time is indexed by years ranging from 1 to 5. The actual years at which these indices indicate is not shown in the data or the documentation. Moreover without further information we assume the sales figurue indicates units sold.

<!-- ----------------------------------------------- --> 
<!-- ------------DATA VISUALISATION------------------ -->
<!-- ----------------------------------------------- --> 
## Data Preperation:
We are dealing with monthly sales data so it may be useful to make a trading day adjustment. Unfortunately, for this, we would require the actual year in which this data took place. However we can make a month length adjustment. The month length adjustment transforms the $y_i$ values of our time series so that they represent the value divided an average month length:
$$w_i=y_i \times \frac{365.25}{12 \times d_i}$$    
Where $w_i,y_i$ are the transformed and untransformed values of the time-series, respectively. $d_i$ is the number of days in month $i$.We are dealing with a 5 year period. This means that either one or two of the years will be leap years. 
Our formula does count for leap years but it requires that the number of days in February, $d_2$, to be $28.25$.
```{r message=FALSE}
d<-c(31,28.25,31,30,31,30,31,31,30,31,30,31)
plastics_transformed<-plastics
for (i in 0:59) 
{
  number<-i%%12 +1
  f<-function(sales)
  {
    output<-sales*(365.25/(12*d[number]))
    return(output)
  }
  
  plastics_transformed[i+1]<-f(plastics_transformed[i+1])
}
plastics_transformed
```


## Data Visualisation:
First we look at the full time series plot of the untransformed data:
```{r message=FALSE}
autoplot(plastics) + ggtitle("Monthly sales of product A for a plastics manufacturer. ") +
  xlab("Time in Years") + ylab("Sales")
```

```{r message=FALSE}
autoplot(plastics_transformed) + ggtitle("Transformed-Monthly sales of product A for a plastics manufacturer. ") +
  xlab("Time in Years") + ylab("Sales")

plot(decompose(plastics_transformed))
```


#### Visual Inspection:
There are a few visual quilities which are evident from the time series plot:

* There is a clear seasonality and general trend.
* The seasonality appears to be a yearly cycle.
* The trend appears to be a gradual increase from year to year.
* The minimum of each year occurs near the beginning of the year.
* The maximum of each year occurs around three-fifths of the way through the year.
* There are evident irregularities in the time series towards the end of year two and near half way through year three.

Further analysis of the time series is needed to understand the data properlly. 



#### Decomposing the Time-Series:
We can decompose the time series in order to get a better understanding of the trend, seasonal and random component.   

First we consider assume the additive model: 
$$Y[t]=T[t]+S[t]+e(t)$$
where $Y$ is the time series, $S$ is the seasonal component and $e$ is the error component; $t$ represents time.
```{r message=FALSE}
plot(decompose(plastics))
```

#### Seasonality Analysis:

